---
title: "Introduction to quakr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to quakr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(lubridate)
library(quakr)
library(ggplot2)
library(leaflet)
```

# Introduction

The quakr package provides tools for analyzing and visualizing earthquake data. This vignette demonstrates the basic functionality of the package, including data loading, cleaning, and visualization.

# Loading Data

While the package is designed to process raw data from the NCEI database (doi: 10.7289/V5TD9V7K), it also includes two datasets: `world` and `southamerica`.\
The `world` dataset contains the whole dataset (as of 05/2024), which has been filtered for complete observations for:

-   geographic and geopolitical location (country, state/city/area, longitude and latitude)

-   magnitude

-   total deaths

This ensures that the the `quakr` functions can be used on this data. Additional column entries were left as-is.

```{r}
#| echo: false
data("world", package = "quakr")
head(world)
```

The `southamerica` dataset is a reduced version containing only complete earthquake observations in South America. Additionally, it has been reduced to contain only minimal information to demonstrate hwo the package works. Let's load and inspect this dataset.

data("world", package = "quakr")

```{r}
#| echo: false
data("southamerica", package = "quakr")
head(southamerica)
```

# Data Cleaning

One of the key functions in **`quakr`** is **`eq_lean_data()`**, which cleans column names and separates country and location information.

```{r}
cleaned_data <- eq_clean_data(world)
head(cleaned_data)
```

# Visualizing Earthquake Data

The package provides several functions for visualizing earthquake data. One of the primary visualization functions is **`geom_timeline`**, which plots earthquake events on a timeline.

### Basic Timeline Plot

```{r}
# Filter the data for Chilean earthquakes since 1970
chile_data <- southamerica %>%
  filter(country == "CHILE" & lubridate::year(date) >= 1970)

# Create a basic timeline plot
ggplot(chile_data, aes(x = date)) +
  geom_timeline(aes(xmin = min(date), xmax = max(date))) +
  theme_minimal() +
  labs(x = "Year",
       size = "Richter Scale Magnitude",
       col = "Total Deaths")
```

### **Adding Labels to the Timeline**

The **`geom_timeline_label`** function can be used to add labels to the timeline plot, making it easier to identify significant events.

```{r}
ggplot(chile_data, aes(x = date)) +
  geom_timeline(aes(xmin = min(date), xmax = max(date), 
                    size = mag, col = total_deaths)) +
  geom_timeline_label(aes(label = location)) +
  theme_minimal() +
  labs(x = "Year",
       size = "Richter Scale Magnitude",
       col = "Total Deaths") +
  theme(legend.position = "bottom")
```

### Using Alternating Labels

Additional to `check_overlap` , which reduces overplotting of labels by blocking the text from being displayed, the `label_dodge` parameter uses alternating vertical orientation of text labels.

```{r}
ggplot(chile_data, aes(x = date)) +
  geom_timeline(aes(xmin = min(date), xmax = max(date), 
                    size = mag, col = total_deaths)) +
  geom_timeline_label(aes(label = location), 
                      check_overlap = TRUE) +
  theme_minimal() +
  labs(x = "Year",
       size = "Richter Scale Magnitude",
       col = "Total Deaths") +
  theme(legend.position = "bottom")
```

```{r}
ggplot(chile_data, aes(x = date)) +
  geom_timeline(aes(xmin = min(date), xmax = max(date), 
                    size = mag, col = total_deaths)) +
  geom_timeline_label(aes(label = location), 
                      check_overlap = TRUE,
                      label_dodge = TRUE) +
  theme_minimal() +
  labs(x = "Year",
       size = "Richter Scale Magnitude",
       col = "Total Deaths") +
  theme(legend.position = "bottom")
```

### **Limiting the Number of Labels**

To avoid overcrowding the plot, you can limit the number of labels displayed based on earthquake magnitude.

```{r}
ggplot(chile_data, aes(x = date)) +
  geom_timeline(aes(xmin = min(date), xmax = max(date),
                    size = mag, col = total_deaths)) +
  geom_timeline_label(aes(label = location, limit = mag), 
                      n_max = 3, 
                      label_dodge = TRUE) +
  theme_minimal() +
  labs(x = "Year",
       size = "Richter Scale Magnitude",
       col = "Total Deaths") +
  theme(legend.position = "bottom")
```

## Interactive Maps

The **`quakr`** package also supports creating interactive maps using **`leaflet`**.

### Basic Map

```{r}
# Adjust point size with the `scaling_factor` parameter
map <- eq_map(cleaned_data, annot_col = "date", scaling_factor = 5)
map
```

### Adding Labels to Maps

You can add more information to the interactive labels via the `eq_create_label()` function. The default information includes location, magnitude and total deaths.

```{r}
chile_data %>% 
  mutate(label = eq_create_label(., defaults = TRUE)) %>% 
  eq_map(annot_col = "label", scaling_factor = 5)
```

`Eq_create_label` allows customization by adding variables to the defaults or replacing the defaults entirely.

```{r}
chile_data %>% 
  mutate(label = eq_create_label(., defaults = TRUE, "date", "mag")) %>% 
  eq_map(annot_col = "label", scaling_factor = 5)
```

```{r}
chile_data %>% 
  mutate(label = eq_create_label(., defaults = FALSE, "date", "mag")) %>% 
  eq_map(annot_col = "label", scaling_factor = 5)
```
